\huge{ \emph{Supplement to} Algorithms For Shaping a Particle Swarm\\ With a Shared Control Input Using Boundary Interaction}

### 2016-09-09 RA-L/experiment.tex
edits, cut to 8 pages

Our experiments are on centimeter-scale hardware systems called \emph{kilobots}.  These allows us to emulate a variety of dynamics, while enabling a high degree of control over robot function, the environment, and data collection. The kilobot, from \citep{Rubenstein2012,rubenstein2014programmable} is a low-cost robot designed for testing collective algorithms with large numbers of robots. It is available as an open source platform or commercially from~\citep{K-Team2015}.  Each robot is approximately 3 cm in diameter, 3 cm tall, and uses two vibration motors to move on a flat surface at speeds up to 1 cm/s.  Each robot has one ambient light sensor that is used to implement \emph{phototaxis},  moving towards a light source. 
In these experiments as shown in Fig.~\ref{fig:setup}, we used $n$=100 kilobots, a 1 m$\times$1 m whiteboard as the workspace, four 30W and four 50W LED floodlights arranged 1.5 m above the plane of the table at the $\{N,NE,E,SE,S,SW,W,NW\}$ vertices of a 6 m square centered on the workspace. The lights were controlled using an Arduino Uno board connected to an 8-relay shield.  Above  the table, an overhead machine vision system tracks the position of the swarm.

\begin{figure}
\begin{center}
	\includegraphics[width=.9\columnwidth]{SetUp.pdf}
\end{center}
\vspace{-1em}
\caption{\label{fig:setup}
Hardware platform:  table with 1$\times$1 m workspace, surrounded by eight  triggered LED floodlights, with an overhead machine vision system.
}
\vspace{-1em}
\end{figure}
\subsection{Hardware Experiment: Position Control of Two Robots}
The walls of the hardware platform have almost infinite friction, due to a laser-cut, slotted border. When a kilobot is steered into the slotted border, they pin themselves to the wall unless the global input directs them away from the wall.  Even though the kilobots are stochastic, this wall friction is sufficient to enable independent control of two kilobots, as shown in Fig.~\ref{fig:storyReal}.

@@

\subsection{Hardware Experiment: Control of Covariance}
To demonstrate covariance control of up to 100 kilobots were placed on the workspace and manually steered with lights, using friction with the boundary walls to vary the covariance from  -4000 to 3000 cm$^2$.  The resulting covariance is plotted in Fig.~\ref{fig:covExperiment}, along with snapshots of the swarm.

>>>>

Our experiments are on centimeter-scale hardware systems called \emph{kilobots}.  These allows us to emulate a variety of dynamics, while enabling a high degree of control over robot function, the environment, and data collection. The kilobot, from \citep{Rubenstein2012,rubenstein2014programmable} is a low-cost robot designed for testing collective algorithms with large numbers of robots. It is available as an open source platform or commercially from~\citep{K-Team2015}.  Each robot is approximately 3 cm in diameter, 3 cm tall, and uses two vibration motors to move on a flat surface at speeds up to 1 cm/s.  Each robot has one ambient light sensor that is used to implement \emph{phototaxis},  moving towards a light source. 
In these experiments as shown in Fig.~\ref{fig:setup}, we used $n$=100 kilobots, a 1 m$\times$1 m whiteboard as the workspace, four 30W and four 50W LED floodlights arranged 1.5 m above the plane of the table at the midpoints and vertices of a 6 m square centered on the workspace. An Arduino Uno connected to an 8-relay shield controlled the lights.  Above  the table, an overhead machine vision system tracked the position of the swarm.

\begin{figure}
\begin{center}
	\includegraphics[width=.9\columnwidth]{SetUp.pdf}
\end{center}
\vspace{-1em}
\caption{\label{fig:setup}
Hardware platform:  table with 1$\times$1 m workspace, surrounded by eight  triggered LED floodlights, with an overhead machine vision system.
}
\vspace{-1em}
\end{figure}
\subsection{Hardware Experiment: Position Control of Two Robots}
The hardware platform walls have almost infinite friction, due to a laser-cut, slotted border. When a kilobot is steered into the slotted border, they pin themselves to the wall unless the global input directs them away from the wall.  Even though the kilobots are stochastic, this wall friction is sufficient to enable independent control of two kilobots, as shown in Fig.~\ref{fig:storyReal}.

@@

\subsection{Hardware Experiment: Control of Covariance}
To demonstrate covariance control, up to 100 kilobots were placed on the workspace and manually steered with lights, using friction with the boundary walls to vary the covariance from  -4000 to 3000 cm$^2$.  The resulting covariance is plotted in Fig.~\ref{fig:covExperiment}, along with snapshots of the swarm.

### 2016-09-09 RA-L/intro.tex
some of Mary's Comments
edits, cut to 8 pages

%One surprising result was that humans that only knew the swarm's mean and covariance completed the task faster that humans who knew the position of every robot~\citep{Becker2013b}. Our previous work focused on a block-pushing task, where a swarm of robots pushed a larger block through a 2D maze. 
Even without obstacles or boundaries, the mean position of the swarm in \eqref{eq:swarmDynamics} is controllable.  By adding rectangular boundary walls, some higher-order moments, such as the swarm's position variance orthogonal to the boundary walls ($\sigma_x$ and $\sigma_y$ for a workspace with axis-aligned walls), are also controllable~\citep{ShahrokhiIROS2015}. 
A limitation is that global control can only compress a swarm orthogonal to obstacles, but navigating through narrow passages often requires control of the variance and the covariance of the swarm.

The paper is arranged as follows. First \S\ref{subsec:FluidInTank} provides analytical position control results in two canonical workspaces with frictionless walls.  These results are limited in the set of shapes that can be generated.  To extend the range of possible shapes, \S \ref{subsec:WallFriction} introduces wall friction to the system model.  We prove that two orthogonal boundaries with high friction are sufficient to arbitrarily position two robots in \S \ref{sec:PostionControl2Robots}, and \S \ref{sec:PostionControlnRobots} extends this to prove a rectangular workspace with high-friction boundaries can position a swarm of $n$ robots arbitrarily within a subset of the workspace.
\S \ref{sec:simulation} describes implementations of both position control algorithms in simulation and  \S \ref{sec:expResults} describes experiments with a hardware setup and up to 100 robots, as shown in Fig.~\ref{fig:IntroPic}. After a review of recent related work \S \ref{sec:RelatedWork}, we end by suggesting directions for future research in \S \ref{sec:conclusion}.
%For controlling $\sigma_{xy}$, we prove that the swarm position covariance $\sigma_{xy}$ is controllable given boundaries with non-zero friction. 
%We then prove that two orthogonal boundaries with high friction are sufficient to arbitrarily position a swarm of $n$ robots. 
%We conclude by designing controllers that efficiently regulate $\sigma_{xy}$.
%This paper
%(1) proves that the swarm position covariance $\sigma_{xy}$ is controllable given boundaries with non-zero friction,  %where do we do this?
%(2) proves that two orthogonal boundaries with high friction are sufficient to arbitrarily position two robots, 
%(3) proves that two orthogonal boundaries with high friction are sufficient to arbitrarily position a swarm of $n$ robots, 
%(4) shows full-state position control with 2 or more robots using  extensive simulations, and
%(5) demonstrate covariance control on our hardware platform with a large number of hardware robots.
%TODO JOURNAL: design controllers that efficiently regulate $\sigma_{xy}$.
%TODO JOURNAL: We will design Lyapunov-inspired controllers for $\sigma_{xy}$ to prove controllability. 
%TODO JOURNAL:  and rank controllability as a function of friction.
% TODO: JOURNAL: and vary wall friction by laser-cutting boundary walls with a variety of profiles. 
%\begin{figure}[t]
%\centering
%\begin{overpic}[width = \columnwidth]{Covariance.jpg}\end{overpic}
%\vspace{-1em}
%\caption{\label{fig:covFriction} Maintaining group cohesion while steering a swarm through an arbitrary maze requires covariance control.
%}\vspace{-1em}
%\end{figure}
% Our paper is organized as follows.  After a discussion of related work in Section \ref{sec:RelatedWork}, we describe our experimental methods for an online human-user experiment in Section \ref{sec:expMethods}.  We report the results of our experiments in Section \ref{sec:expResults}, discuss the lessons learned in Section \ref{sec:discussion}, and end with concluding remarks in Section \ref{sec:conclusion}.

>>>>

%One surprising result was that humans that only knew the swarm's mean and covariance completed the task faster that humans who knew the position of every robot~\citep{Becker2013b}. Our previous work focused on a block-pushing task, where a swarm of robots pushed a larger block through a 2D maze. 
Even without obstacles or boundaries, the mean position of the swarm in \eqref{eq:swarmDynamics} is controllable.  By adding rectangular boundary walls, some higher-order moments, such as the swarm's position variance orthogonal to the boundary walls ($\sigma_x$ and $\sigma_y$ for a workspace with axis-aligned walls), are also controllable~\citep{ShahrokhiIROS2015}. 
%A limitation is that global control can only compress a swarm orthogonal to obstacles, but navigating through narrow passages often requires control of the variance and the covariance of the swarm.
These techniques could only compress a swarm orthogonal to obstacles.
Swarm shape control is necessary for navigation through narrow passages, for self-assembly, and for manipulating objects.

The paper is arranged as follows. First \S\ref{subsec:FluidInTank} provides analytical position control results in two canonical workspaces with frictionless walls.  These results are limited in the set of shapes that can be generated.  To extend the range of possible shapes, \S \ref{subsec:WallFriction} introduces general models for boundary interaction.  We prove that two orthogonal boundaries with high friction are sufficient to arbitrarily position two robots in \S \ref{sec:PostionControl2Robots}, and \S \ref{sec:PostionControlnRobots} extends this to prove a rectangular workspace with high-friction boundaries can position a swarm of $n$ robots arbitrarily within a subset of the workspace.
Both algorithms require robots with low process noise and require time proportional to the number of robots, so \S \ref{subsec:ClosedLoopCovarianceControl} introduces a robust and efficient technique to control swarm covariance.
\S \ref{sec:simulation} describes implementations of the algorithms in simulation and  \S \ref{sec:expResults} describes hardware experiments with up to 100 robots, as shown in Fig.~\ref{fig:IntroPic}. After a review of recent related work in \S \ref{sec:RelatedWork}, we end with directions for future research in \S \ref{sec:conclusion}.
%For controlling $\sigma_{xy}$, we prove that the swarm position covariance $\sigma_{xy}$ is controllable given boundaries with non-zero friction. 
%We then prove that two orthogonal boundaries with high friction are sufficient to arbitrarily position a swarm of $n$ robots. 
%We conclude by designing controllers that efficiently regulate $\sigma_{xy}$.
%This paper
%(1) proves that the swarm position covariance $\sigma_{xy}$ is controllable given boundaries with non-zero friction,  %where do we do this?
%(2) proves that two orthogonal boundaries with high friction are sufficient to arbitrarily position two robots, 
%(3) proves that two orthogonal boundaries with high friction are sufficient to arbitrarily position a swarm of $n$ robots, 
%(4) shows full-state position control with 2 or more robots using  extensive simulations, and
%(5) demonstrate covariance control on our hardware platform with a large number of hardware robots.
%TODO JOURNAL: design controllers that efficiently regulate $\sigma_{xy}$.
%TODO JOURNAL: We will design Lyapunov-inspired controllers for $\sigma_{xy}$ to prove controllability. 
%TODO JOURNAL:  and rank controllability as a function of friction.
% TODO: JOURNAL: and vary wall friction by laser-cutting boundary walls with a variety of profiles. 
%\begin{figure}[t]
%\centering
%\begin{overpic}[width = \columnwidth]{Covariance.jpg}\end{overpic}
%\vspace{-1em}
%\caption{\label{fig:covFriction} Maintaining group cohesion while steering a swarm through an arbitrary maze requires covariance control.
%}\vspace{-1em}
%\end{figure}
% Our paper is organized as follows.  After a discussion of related work in Section \ref{sec:RelatedWork}, we describe our experimental methods for an online human-user experiment in Section \ref{sec:expMethods}.  We report the results of our experiments in Section \ref{sec:expResults}, discuss the lessons learned in Section \ref{sec:discussion}, and end with concluding remarks in Section \ref{sec:conclusion}.

### 2016-09-09 RA-L/simulation.tex
Edits made to the paper
edits, cut to 8 pages

Two simulations were implemented using wall-friction for position control.  The first controls the position of two robots, the second controls the position of $n$ robots.  All code is available online at \citep{Zhao2016mathematica,Zhao2016mathematicaSquare}.

@@

Algorithms \ref{alg:PosControl2Robots}, \ref{alg:XControl}, %\ref{alg:YControl}, 
were implemented in Mathematica using point robots (radius = $0$).  Fig.~\ref{fig:shapeControlMathematica1}  shows  this algorithm for two configurations. 
Robot initial positions are shown by a crosshair, and final positions by a circled crosshair.  Dashed lines show the shortest route if robots could be controlled independently.  The path given by  Alg.\ \ref{alg:PosControl2Robots} is shown with solid arrows.
Each row has five snapshots taken every quarter second. For the sake of brevity axis-aligned moves were replaced with oblique moves that combine two moves simultaneously. 
 $\Delta r_x$ is adjusted to $\Delta e_x$ in the second snapshot at $t = 0.25$. 
 The following frames  adjust $\Delta r_y$ to $\Delta e_y$. 
 $\Delta r_y$ is corrected by $t = 0.75$. 
 Finally, the algorithm moves the robots to their corresponding destinations.

@@

\subsection{Position Control of $n$ Robots}
Alg. \ref{alg:PosControlNRobots}  was simulated in {\sc Matlab} using square block robots with unity width. Code is available at \citep{Arun2015}.
Simulation results are shown in Fig.~\ref{fig:4diagramsplots.pdf} for arrangements with an increasing number of robots,  $n$= [8, 46, 130, 390, 862]. 
The distance moved grows quadratically with the number of robots $n$. A best-fit line $210 n^2 + 1200n-10,000$ is overlaid by the data..

@@

\subsection{Efficient Control of Covariance}
%A set of simulations were conducted to demonstrate the importance of boundary friction.  
Random disturbances impair the performance of Alg. \ref{alg:PosControl2Robots} and Alg. \ref{alg:PosControlNRobots}. Still, we are able to control covariance of the swarm. This section demonstrates simulations of controlling covariance of the swarm.
These simulations use  the 2D physics engine Box2D, by \citep{catto2010box2d}.
 144 disc-shaped robots were controlled by an open-loop control input as illustrated in Fig. \ref{fig:SimCovarianceFuncFrictionOpenLoop}.  All robots had  the same initial conditions, but in four tests the boundary friction was $F_f = \{0,1/3 F, 2/3F, F\}$.
 Without friction, covariance has minimal variation.  As friction increases, the covariance can be manipulated to greater degrees.

>>>>

Two simulations were implemented using wall-friction for position control.  The first controls the position of two robots, the second controls the position of $n$ robots.

@@

Algorithms \ref{alg:PosControl2Robots} and \ref{alg:XControl}, %\ref{alg:YControl}, 
were implemented in Mathematica using point robots (radius = $0$).  Fig.~\ref{fig:shapeControlMathematica1}  shows  an implementation of this algorithm. 
Robot initial positions are shown by a crosshair, and final positions by a circled crosshair.  Dashed lines show the shortest route if robots could be controlled independently.  The path given by  Alg.\ \ref{alg:PosControl2Robots} is shown with solid arrows.
Each row has five snapshots taken every quarter second. For the sake of brevity axis-aligned moves were replaced with oblique moves that combine two moves simultaneously. 
 $\Delta r_x$ is adjusted to $\Delta e_x$ in the second snapshot at $t = 0.25$. 
 The following frames  adjust $\Delta r_y$ to $\Delta e_y$. 
 $\Delta r_y$ is corrected by $t = 0.75$. 
 Finally, the algorithm moves the robots to their corresponding destinations.

@@

\subsection{Position Control of $n$ Robots}
Alg. \ref{alg:PosControlNRobots}  was simulated in {\sc Matlab} using square block robots with unity width. Code is available at \citep{Arun2015}.
Simulation results are shown in Fig.~\ref{fig:4diagramsplots.pdf} for arrangements with an increasing number of robots,  $n$= [8, 46, 130, 390, 862]. 
The distance moved grows quadratically with the number of robots $n$. A best-fit line $210 n^2 + 1200n-10,000$ is overlaid by the data.

@@

\subsection{Efficient Control of Covariance}
%A set of simulations were conducted to demonstrate the importance of boundary friction.  
Random disturbances impair the performance of Alg. \ref{alg:PosControl2Robots} and Alg. \ref{alg:PosControlNRobots}. Due to the central limit theorem, we can control the covariance of large swarms. This section demonstrates controlling swarm covariance in simulation, using the 2D physics engine Box2D~\citep{catto2010box2d}.
 144 disc-shaped robots were controlled by an open-loop control input as illustrated in Fig. \ref{fig:SimCovarianceFuncFrictionOpenLoop}.  All robots had  the same initial conditions, but in four tests the boundary friction was $F_f = \{0,1/3 F, 2/3F, F\}$.
 Without friction, covariance has minimal variation.  As friction increases, the covariance can be manipulated to greater degrees.

### 2016-09-09 RA-L/theory.tex
edits, cut to 8 pages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sec:theory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=\linewidth]{SquareFill.pdf} 
\vspace{-1em}
\caption{Pushing the swarm against a square boundary wall allows limited control of the shape of the swarm, as a function of swarm area $A$ and the commanded movement direction $\beta$. Left plot shows locus of possible mean positions for five values of $A$.  The locus morphs from a square to a circle as $A$ increases.  The covariance ellipse for each $A$ is shown with a dashed line. Center shows two corresponding arrangements of kilobots.  At right is $\bar{x}(A), \sigma_{xy}(A), \sigma_x^2(A),$ and $\rho(A)$ for a range of $\beta$ values. See online interactive demonstration at \citep{Zhao2016mathematicaSquare}.}
\label{fig:SquareFill}
\end{center}
\end{figure*} 
\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=\linewidth]{CircleFill.pdf} 
\vspace{-1em}
\caption{Pushing the swarm against a circular boundary wall allows limited control of the shape of the swarm, as a function of the fill level $h$ and the commanded movement direction $\beta$. Left plot shows locus of possible mean positions for four values of $h$. The locus of possible mean positions are concentric circles. See online interactive demonstration at \citep{Zhao2016mathematica}.}  
\label{fig:CircleFill}
\end{center}
\end{figure*}

\subsection{Using Boundaries: Fluid Settling In a Tank}\label{subsec:FluidInTank}
One method to control a swarm's shape in a bounded workspace is to simply push in a given direction until the swarm conforms to the boundary.
\paragraph{Square workplace}
%The formulas for means $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ are as follows: 
We first examine the mean $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ of a very large swarm of robots as they move inside a square workplace under the influence of gravity pointing in the direction $\beta$. The swarm is large, but the robots are small in comparison, and together occupy a constant area $A$. Under a global input such as gravity, they flow like water, moving to a side of the workplace and forming a polygonal shape, as shown in Fig.~\ref{fig:SquareFill}.

The range for the global input angle $\beta $ is [0,2$\pi $). In this range, the swarm assumes eight different polygonal shapes. Therse shapes alternate between triangles and trapezoids when the area $A$$<$1/2, and between squares with one corner removed and trapezoids when $A$$>$1/2.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Circular workplace}
Though rectangular boundaries are common in artificial workspaces, biological workspaces are usually rounded.
Similar calculations can be made for a circular workspace.  The workspace is a circle centered at (0,0) with radius 1 and thus area $\pi$.
For notational simplicity, the swarm is parameterized by the global control input signal $\beta$ and the fill-level $h$.  
Under a global input, the robot swarm fills the region under a chord with area
\begin{align}
A(h) = \arccos(1-h)-(1-h) \sqrt{(2-h) h}.
\end{align}
For a circular workspace, the locus of mean positions are aligned with $\beta$ and the mean position is at radius $r(h)$ from the center:
\begin{align}
r(h) = \frac{2 (-(h-2) h)^{3/2}}{3 \left(\sqrt{-(h-2) h} (h-1)+\arccos(1-h)\right)}
\end{align}
Variance $\sigma^2_x(\beta,h)$ is maximized at $\beta = \pi/2+n \pi$ and $h\approx1.43$, while covariance is maximized at $\beta = \pi3/4+n \pi$ and $h\approx0.92.$ For small $h$ values, correlation approaches $\pm1$. Results are displayed in Fig.~\ref{fig:CircleFill}.

@@

Let the control input be a vector force $\vec{F}$ with magnitude $F$ and orientation $\theta$ with respect to a line perpendicular to and into the nearest boundary. $N$ is the normal or perpendicular force between the robot and the boundary. The force of friction $F_f$ is nonzero if the robot is in contact with the boundary and  $|\theta| < \pi/2$. The resulting net force on the robot, $F_{\text{\emph{forward}}}$, is aligned with the wall and given by

\begin{align}
F_{\text{\emph{forward}}} &=  F \sin(\theta) - F_f  \nonumber \\
\text{where }  F_f &= \begin{cases}  \mu_f N, &  \mu_f N < F \sin(\theta)  \label{eq:frictionmodel}  \\
F \sin(\theta), & \text{else} \end{cases} \\ %\sign(F \sin(\theta) ) \cdot  \max(0, | F sin\theta |- |F_f|)
\text{and } N &= F \cos(\theta) \nonumber
\end{align}
 Fig.~\ref{fig:friction} shows the resultant forces on two robots when one is touching a wall. As illustrated, both experience different net forces although each receives the same inputs.
  For ease of analysis, the following algorithms assume $\mu_f$ is infinite and robots touching the wall are prevented from sliding along the wall.
This means that if one robot is touching the wall and another robot is free, the touching robot will not move when the control input is parallel or into the wall. 
There are many alternate models of friction that also break control symmetry. Fig.~\ref{fig:friction}c shows fluid flow along a boundary.  Fluid in the free-flow region moves uniformly, but flow decreases to zero in the boundary layer.  
\begin{align}
u(y) = u_0 [1- \frac{(y-h)^2}{h^2} = u_0 \frac{y}{h} [2- \frac{y}{h}] \label{eq:boundarylayerflow}
\end{align}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sec:theory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=\linewidth]{SquareFill.pdf} 
\vspace{-1em}
\caption{Pushing the swarm against a square boundary wall allows limited control of the shape of the swarm, as a function of swarm area $A$ and the commanded movement direction $\beta$. Left plot shows locus of possible mean positions for five values of $A$.  The locus morphs from a square to a circle as $A$ increases.  The covariance ellipse for each $A$ is shown with a dashed line. Center shows two corresponding arrangements of kilobots.  At right is $\bar{x}(A), \sigma_{xy}(A), \sigma_x^2(A),$ and $\rho(A)$ for a range of $\beta$ values. See online interactive demonstration at \citep{Zhao2016mathematicaSquare}.}
\label{fig:SquareFill}
\end{center}
\end{figure*} 
\begin{figure*}[!htb]
\begin{center}
\vspace{-1em}
\includegraphics[width=\linewidth]{CircleFill.pdf} 
\vspace{-1em}
\caption{Pushing the swarm against a circular boundary wall allows limited control of the shape of the swarm, as a function of the fill level $h$ and the commanded movement direction $\beta$. Left plot shows locus of possible mean positions for four values of $h$. The locus of possible mean positions are concentric circles. See online interactive demonstration at \citep{Zhao2016mathematica}.}  
\label{fig:CircleFill}
\end{center}
\end{figure*}

\subsection{Using Boundaries: Fluid Settling In a Tank}\label{subsec:FluidInTank}
One method to control a swarm's shape in a bounded workspace is to simply push in a given direction until the swarm conforms to the boundary.
\paragraph{Square workplace}
%The formulas for means $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ are as follows: 
We first examine the mean $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ of a very large swarm of robots as they move inside a square workplace under the influence of gravity pointing in the direction $\beta$. The swarm is large, but the robots are small in comparison, and together occupy a constant area $A$. Under a global input such as gravity, they flow like water, moving to a side of the workplace and forming a polygonal shape, as shown in Fig.~\ref{fig:SquareFill},  see also \citep{Zhao2016mathematicaSquare}.

The range for the global input angle $\beta $ is [0,2$\pi $). In this range, the swarm assumes eight different polygonal shapes. 
These shapes alternate between triangles and trapezoids when the area $A$$<$1/2, and between squares with one corner removed and trapezoids when $A$$>$1/2.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Circular workplace}
Though rectangular boundaries are common in artificial workspaces, biological workspaces are usually rounded.
Similar calculations can be made for a circular workspace.  The workspace is a circle centered at (0,0) with radius 1 and thus area $\pi$.
For notational simplicity, the swarm is parameterized by the global control input signal $\beta$ and the fill-level $h$.  
Under a global input, the robot swarm fills the region under a chord with area
\begin{align}
A(h) = \arccos(1-h)-(1-h) \sqrt{(2-h) h}.
\end{align}
For a circular workspace, the locus of mean positions are aligned with $\beta$ and the mean position is at radius $r(h)$ from the center:
\begin{align}
r(h) = \frac{2 (-(h-2) h)^{3/2}}{3 \left(\sqrt{-(h-2) h} (h-1)+\arccos(1-h)\right)}
\end{align}
Variance $\sigma^2_x(\beta,h)$ is maximized at $\beta = \pi/2+n \pi$ and $h\approx1.43$, while covariance is maximized at $\beta = \pi3/4+n \pi$ and $h\approx0.92.$ For small $h$ values, correlation approaches $\pm1$. Results are displayed in Fig.~\ref{fig:CircleFill}, see also \citep{Zhao2016mathematica}.

@@

Let the control input be a vector force $\vec{F}$ with magnitude $F$ and orientation $\theta$ with respect to a line perpendicular to and into the nearest boundary. $N$ is the normal or perpendicular force between the robot and the boundary. The force of friction $F_f$ is nonzero if the robot is in contact with the boundary and  $|\theta| < \pi/2$. The resulting net force on the robot, $F_{\text{\emph{forward}}}$, is aligned with the wall and given by
\begin{align}
F_{\text{\emph{forward}}} &=  F \sin(\theta) - F_f  \nonumber \\
\text{where }  F_f &= \begin{cases}  \mu_f N, &  \mu_f N < F \sin(\theta)  \label{eq:frictionmodel}  \\
F \sin(\theta), & \text{else} \end{cases} \\ %\sign(F \sin(\theta) ) \cdot  \max(0, | F sin\theta |- |F_f|)
\text{and } N &= F \cos(\theta) \nonumber
\end{align}
 Fig.~\ref{fig:friction} shows the resultant forces on two robots when one is touching a wall. 
Though each receives the same inputs,  they experience different net forces.
  For ease of analysis, the following algorithms assume $\mu_f$ is infinite and robots touching the wall are prevented from sliding along the wall.
This means that if one robot is touching the wall and another robot is free, the touching robot will not move when the control input is parallel or into the wall. 
There are many alternate models of friction that also break control symmetry. Fig.~\ref{fig:friction}c shows fluid flow along a boundary.  Fluid in the free-flow region moves uniformly, but flow decreases to zero in the boundary layer.  
\begin{align}
u(y) = u_0 [1- \frac{(y-h)^2}{h^2} = u_0 \frac{y}{h} [2- \frac{y}{h}] \label{eq:boundarylayerflow}
\end{align}
